Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-05-01 sources/bmc_template_submission/bmc_article.tex
last edits to captions

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to simultaneously study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across biological databases.
Physcraper's main goal is to use its interoperability framework to build upon published alignments and phylogenies
to extend existing phylogenetic inferences with more data and taxa, which improve our
phylogenetic reconstructions \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
and biogeographic analyses \cite{kayaalp2017back},
and help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Physcraper updates a starting single locus alignment and corresponding phylogeny
with public DNA data from GenBank \cite{benson2000genbank}, and links tip taxon labels in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}, streamlining connections between phylogenies and biological data.
Physcraper links tip taxon labels in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}, eefectively streamlining connections between phylogenies and biological data.

@@

BLAST is performed using the blastn algorithm \cite{camacho2009blast}
implemented in BioPython 1.71 \cite{cock2009biopython} NCBIWWW module \cite{ncbiwww}
modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against the GenBank database.
Matching sequences are filtered and excluded from the analysis if they 1) are not in the search taxon;
2) have an e-value above the cutoff (default to 0.00001); 3) fall outside a minimum
and maximum sequence length threshold, defined as a proportion of the average sequence length without
gaps of all sequences in the input alignment (default values of 80\% and 120\%,
respectively); 4) or, if they are either identical to or shorter than an existing
sequence in the input alignment and they represent the same taxon in NCBI's or
OpenTree's unififed taxonomy.
By default, an arbitrary maximum number of 5 sequences per taxon are chosen at random
from the set of matching sequences that passed the filtering step.
(default to 5).

@@

Physcraper generates gene trees, which individually do not capture the full complexity
of species' evolutionary history \cite{song2012resolving}. In addition, single gene
phylogenies with very high numbers of taxa may lack sufficient signal
for accurate phylogenetic resolution \cite{morel_phylogenetic_2020}.
The Physcraper workflow avoids this challenge by focusing on ingroup
taxa of an existing phylogeny, using markers that have been assessed and proven
appropriate for that phylogenetic scope in past publications.
Also, Physcraper thins alignments by removing sequences identical to original
and newly added sequences, and by setting a maximum number of sequences per taxon.
Nonetheless, it is incumbent on users to assess their final inference with respect
to statistical support and biological plausibility.

@@

\begin{figure}[h!]
  \caption{A) Phylogeny updated with Physcraper using a starting phylogeny and an alignment from
  \cite{gottlieb2005molecular} (Gottlieb2005 data in text).
  Tips in original alignment and new tips added with Physcraper are depicted in
  black and red, respectively. Physcraper obtained sequences from the GenBank
  database via local BLAST of all sequences from the Gottlieb2005 internal transcribed
  spacer DNA region (ITS) original alignment,
  filtered them following criteria described in section ``DNA sequence search and
  filtering", aligned them to the original alignment using MUSCLE, and performed a
  phylogenetic reconstruction using RAxML, with 100 bootstraps. B) Results of conflict
  analysis performed using OpenTree's conflict tool \cite{redelings2017supertree}.
  The Physcraper updated Gottlieb2005 phylogeny in (A) was compared to an \textit{Ilex}
  OpenTree synthetic subtree v. 12.3 \cite{ott727571} constructed using taxonomy of the genus
  as backbone and resolving branches based on phylogenetic data from the original
  Gottlieb2005 phylogeny.
}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to simultaneously study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across biological databases.
Physcraper's main goal is to use its interoperability framework to build upon published
alignments and phylogenies
to extend existing phylogenetic inferences with sequence data available in public
DNA databases for more taxa. Increased taxonomic sampling improves
phylogenetic reconstructions \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
and biogeographic analyses \cite{kayaalp2017back},
and help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
% Physcraper updates the starting single locus alignment and corresponding phylogeny
% with public DNA data from GenBank \cite{benson2000genbank}, and links tip taxon
% labels in the updated trees to a unified, interoperable
% taxonomic resource \cite{rees2017automated}, streamlining connections between phylogenies and biological data.
Physcraper links tip taxon labels in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}, effectively streamlining connections between phylogenies and biological data.

@@

BLAST is performed using the blastn algorithm \cite{camacho2009blast}
implemented in BioPython 1.71 \cite{cock2009biopython} NCBIWWW module \cite{ncbiwww}
modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against the GenBank database.
Matching sequences are filtered and excluded from the analysis if they 1) are not in the search taxon;
2) have an e-value above the cutoff (default to 0.00001); 3) fall outside a minimum
and maximum sequence length threshold, defined as a proportion of the average sequence length without
gaps of all sequences in the input alignment (default values of 80\% and 120\%,
respectively); 4) or, if they are either identical to or shorter than an existing
sequence in the input alignment and they represent the same taxon in NCBI's or
OpenTree's unififed taxonomy.
By default, an arbitrary maximum number of 5 sequences per taxon are chosen at random
from the set of matching sequences that passed the filtering step.

@@

Physcraper generates gene trees, which individually do not capture the full complexity
of species' evolutionary history \cite{song2012resolving}. In addition, single gene
phylogenies with very high numbers of taxa may lack sufficient signal
for accurate phylogenetic resolution \cite{morel_phylogenetic_2020}.
The Physcraper workflow avoids this challenge by focusing on ingroup
taxa of an existing phylogeny, using markers that have been assessed and proven
appropriate for that phylogenetic scope in past publications.
Also, Physcraper thins alignments by removing sequences identical to original
and newly added sequences, and by setting a maximum number of sequences per taxon.
Nonetheless, it is incumbent on users to assess their final inference with respect
to statistical support and biological plausibility.
% Add citation of good things about thining alignments

@@

\begin{figure}[h!]
  \caption{A) Phylogeny updated with Physcraper using a starting phylogeny and an alignment from
  \cite{gottlieb2005molecular} (Gottlieb2005 data in text).
  Tips in original alignment and new tips added with Physcraper are depicted in
  black and red, respectively. Physcraper obtained sequences from the GenBank
  database via local BLAST of all sequences from the Gottlieb2005 internal transcribed
  spacer DNA region (ITS) original alignment,
  filtered them following criteria described in section ``DNA sequence search and
  filtering", aligned them to the original alignment using MUSCLE, and performed a
  phylogenetic reconstruction using RAxML, with 100 bootstraps. B) Results of conflict
  analysis performed using OpenTree's conflict tool \cite{redelings2017supertree}.
  The Physcraper updated Gottlieb2005 phylogeny in (A) was compared to an \textit{Ilex}
  OpenTree synthetic subtree v. 12.3 \cite{ott727571} constructed using taxonomy of the genus
  as backbone and resolving branches based on phylogenetic data from the original
  Gottlieb2005 phylogeny. Green branches are resolved by the updated phylogeny,
  blue branches are in agreement between the updated phylogeny and the syntehtic subtree, and
  orange branches are in conflict between the two.
}
\end{figure}
